Analysis-ready VCF at Biobank-scale using Zarr

### 2024-05-07 paper.tex
Tweaks

VCF is at its core an encoding of the genotype matrix $G$, where each entry
encodes the observed genotypes for a given sample at a give variant site,
interleaved with per-variant information
and other matrices (e.g., GQ or AD).
The data is largely numerical and of a fixed dimension, and is therefore a 
natural mapping to an array-oriented approach [PICK UP HERE]

[TODO intro passage here on array-based or ``tensor'' storage. Separate
out the different interleaved matrices with the VCF, and store them
in regular chunked grids.]

We propose the VCF Zarr specification which maps the 
VCF data model into a columnar layout using the
widely-used Zarr format (Fig~\ref{fig-data-model}). 
Briefly, Zarr is a format 
% FIXME two different uses of $n$ here in a paragraph
for storing $n$-dimensional arrays of a fixed datatype, stored as 
compressed chunks. In the VCF Zarr specification, 
each field in a VCF is mapped to an array, which is compressed 
and stored individually allowing for efficient retrieval. 
In particular, call-level data is stored as $m \times n$ arrays
(for $m$ sites and $n$ samples), allowing for efficient 
retrieval of subsets of those fields along both the 
variants and samples axis.
Note that this approach to storing higher dimensional data
sets Zarr apart from more general columnar formats 
such as Parquet [TODO more details and citations?].
See the Methods for more detail on Zarr and the VCF Zarr
specification.

>>>>

VCF is at its core an encoding of the genotype matrix where each entry
describes the observed genotypes for a given sample at a give variant site,
interleaved with per-variant information
and other matrices (e.g., the GQ or AD fields).
The data is largely numerical and each field is of a fixed dimension, 
and is therefore a natural mapping to array-oriented 
or ``tensor'' storage.
We propose the VCF Zarr specification which maps the 
VCF data model into an array-oriented layout using the
widely-used Zarr format (Fig~\ref{fig-data-model}). 
Briefly, Zarr is a format 
% FIXME two different uses of $n$ here in a paragraph
for storing $n$-dimensional arrays of a fixed datatype, stored as 
compressed chunks. In the VCF Zarr specification, 
each field in a VCF is mapped to an array, which is compressed 
and stored individually allowing for efficient retrieval. 
In particular, call-level data is stored as $m \times n$ arrays
(for $m$ sites and $n$ samples), allowing for efficient 
retrieval of subsets of those fields along both the 
variants and samples axis.
See the Methods for more detail on Zarr and the VCF Zarr
specification.
